Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-26 WSSSPE2_report/WSSSPE2_review.tex

\katznote{need something about subgroup A here}

Subgroup B discussed four papers and had six participants. Because of the nature
of the papers, training emerged as a common theme. However the conversation was
wide-ranging, including incentives, reproducibility, and funding to promote
sustainability.

@@

@@

>>>>

Subgroup A consisted of ten participants who discussed papers surrounding training
and successful community software initiatives.
Subgroup B discussed four papers and had six participants. Because of the nature
of the papers, training emerged as a common theme. However the conversation was
wide-ranging, including incentives, reproducibility, and funding to promote
sustainability.

@@

\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}

Group discussion started with a position statement by each person surrounding what they had
learned form their software development experiences and how those lessons might be translated
into actionable outcomes.  Participants came from a range of backgrounds, and represented
multiple software and training initiatives, including Software Carpentry, Data 
Carpentry, Open Science for Synthesis, the Community Surface Dynamics Modeling Systems group, 
ROpenSci, DataONE, the HDF Group, and others. Some of these software experiences were focused on development
of new products for use in the sciences (e.g., ROpenSci, CSDMS), and these recounted the 
difficulties of engaging with disciplinary scientists in writing software.  Software was clearly
utilized broadly across the various science disciplines represented, and it was developed within those
disciplines as well.  Some researchers created software for statistical analysis and modeling, while
others used it to control instrumentation, collate data across networks, collect information from
respondents, and many other uses.  There was broad consensus that, within the disciplines represented,
formal training in any type of software development or engineering was rare among the practitioners;
most are self-taught, and develop software to get another job done. Any ancillary utility of the software
outside of the specific science target was generally unplanned and few researchers would want to invest more
time to make their own software more re-usable.

Despite agreement that this body of disciplinary software needed improvement, there was general agreement that
such improvement needed to be understood in terms of the scientific productivity that could be acheived. A
software maturity model is needed for science software, but it needs to be introduced in a way that fits
in with the culture of science, which largely thinks of software as a tool, rather than a product itself. 
The group was in general consensus that more widespread training in software practice is needed within the
domain sciences, and several of the participants were involved in efforts along these lines.  Participants felt
that projects that build a sense of community via training in software and technical practices would have the
most success in changing practices in that community, but that there was a need for a managed introduction of
these practices. Participants also recognized that these could not be one-time, one-off training opportunities,
as software and technologies change rapidly over time.  For example, while today Software Carpentry is focused
on teaching version control via Git, there has been a rapid evolution from RCS to CVS to SVN to Git over a short
time frame, and thus communities should expect to need to train for adaptibility and a changing technology tool
chain, rather than assume these technologoes stay fixed.  Thus, although short term training that introduces
specific tools was considered highly valuable, these trainins were also not considered sufficient to engender the changes
in software practice that were deemed neccessary.  Combining the need for changes in practice to be introduced
incrementally with the need to to minimize the divergence of training from direct science goals and the 
difficulties of training for a rapidly evolving technology space, participants concluded that multiple training
efforts that targeted different parts of the spectrum were needed.  Short-term courses introducing immediately
useful skills needs to be offered alongside more in-depth courses on software engineering and practice that allow
students to adapt to a changing landscape.

Finally, after agreeing that these complementary training models were needed, the group discussed sustainability of
training, and how the leading groups in this space are teaching only a small fraction of the community that
needs and wants training.  Most graduate programs in the sciences do not currently incorporate these appraoches
in their graduate curricula, although there are an increasing number of quantitatively oriented courses around
analysis and modeling.  These still, however, generally omit ngineering practices such as version control, unit
and regression testing, and software modularity and abstraction, often because the instructors themselves in the 
domain sciences are not familiar with these techniques.  Thus, students who are being trained in these appraches
are doing so through short 2-3 day training workshops such as Software and Data Carpentry, rather than through 
semester-long graduate education courses at their University, which tend to focus on statistical and modeling 
techniques.  Hybrid programs like the three week Open Science for Synthesis (OSS) training that combine the three 
(science, quantitiative techniques, and software engineering) into an holistic course serve part of the need but 
reach only a few researchers at this time.  Thus, the group concluded that significant strides needed to be made
in coordinating these trainings so that they are complementary, on increasing their size and scope, and on
better integrating them directly into graduate education programs.

@@

Three main actions were identified by Subgroup A that would be of interest to participants and that
would benefit the research community.  These focused on the desire to amplify the 
current community efforts in training and software engineering by conecting the current training initiatives.
( e.g., software/data carpentry, rOpenSci, OSS) of different durations at appropriate stages. Generally it 
was felt that a modicum of interaction between short term workshops (SwC, DC), medium term trainings (e.g., OSS),
and longer-term courses (e.g., BIDS Data Science) would benefit from coordinating curriculum,
discussing and aligning prerequisites, and coordinating timing of courses.  A training
coordination effort would go a long ways towards amplifying the value of the individual
efforts and make them all more effective.

\begin{itemize}
    \item Action 1: Create a roadmap of research software training initiatives. Such a roadmap would
        provide a taxonomy of training opportunities: what they deliver, and what is needed to know 
        going in (prerequisites). It would also show which recommended roadmap actions would have a clear 
        and immediate payoff, and which will have long-term payoffs.  The training roadmap would emphasize 
        time savings and efficiency gains to be had from each training.

    \item Action 2: Build a report card characterizing use of best practices in sceintific software. Generally,
        people felt that researchers would be very willing to migrate practices if they could identify
        where they needed to improve. Such a report card would ask simple questions to charactize use of best
        practices in science software, and could be structured similarly to Joel Spolsky's 'Software Maturity'
        questionairre (http://www.joelonsoftware.com/articles/fog0000000043.html).  The survey would create a 
        report card that shows areas where a project could improve, and then link those areas to specific 
        training offerings from the training taxonomy from Action 1.

    \item Action 3: Create science software review forum.  It was generally acknowledged that a little code review
        can have a tremendous impact on the quality of software in a project, but that sites for science code
        review are lacking.  While people can ask questions on the Stack Exchange sites, they are generally not open
        to questions of style, approach, or appropriateness, as they try to avoid subjective commentary.   Instead, 
        we need a site where code gets discussed/summarized/described (in small bites) by the science software community.
        The target audience would be graduate and undergraduate students in the sciences, and there would need to be
        mechanisms to keep the review positive and constructive, and not get pedantic or judgemental.  
        This could be tied to a software registry (such as the nascent GeoSoft project), or to language repositories like 
        R's CRAN repository, and could lead to the report card discussed in Action 2.
\end{itemize}
